Behavioral Guardrails for Dynamic LLM Persona

### 2025-05-11 alife_manuscript/manuscript.tex
removed old repo link

This work presents a method for aligning LLMs using instruction fine-tuning and preference optimization, creating reliable safeguards while maintaining flexible persona instructions. We carefully prepare training data for generalization and outline a framework where a single guardrail definition aligns the model. Meta-prompts automate synthetic dataset creation for generating Low-Rank Adaptation (LoRA) weights, synonymously referred to as guardrail adapters.

In addition, as part of the framework, we assess the guardrail efficacy, and find a consistent and reliable adherence to the desired behaviors when the guardrail adapters are applied, without a sizeable impact on the natural, safe, conversations.

@@

We found that a straightforward linear merging of the LoRA adapters yields unexpected behavior, resulting in nearly no instances of coherent and guardrail adhering responses when using the same test datasets. 
Instead, we perform a singular value decomposition on the weighted combination of the adapters, truncate to the rank used for the LoRA adapter,
(see~\cite{code}). The resulting merged adapter is able to detect and resolve triggering conversations from all three test sets. 
We retain outputs that adhere to the guardrail definitions with a lower bound of \minGuardrailMergedEfficiency\% accuracy. We summarize our findings in the Tab.~\ref{tab:guardrail-effectiveness}, where merged adapter refers to the adapter resulting from the SVD-based merging of the LoRA adapters.

>>>>

This work presents a method for aligning LLMs using instruction fine-tuning and preference optimization, creating reliable safeguards while maintaining flexible persona instructions. We carefully prepare training data for generalization and outline a framework where a single guardrail definition aligns the model. Meta-prompts automate synthetic dataset creation for generating Low-Rank Adaptation (LoRA) weights, synonymously referred to as guardrail adapters.  
In addition, as part of the framework, we assess the guardrail efficacy, and find a consistent and reliable adherence to the desired behaviors when the guardrail adapters are applied, without a sizeable impact on the natural, safe, conversations.

@@

We found that a straightforward linear merging of the LoRA adapters yields unexpected behavior, resulting in nearly no instances of coherent and guardrail adhering responses when using the same test datasets. 
Instead, we perform a singular value decomposition on the weighted combination of the adapters, truncate to the rank $r=16$, used for the LoRA adapter. The resulting merged adapter is able to detect and resolve triggering conversations from all three test sets. 
We retain outputs that adhere to the guardrail definitions with a lower bound of \minGuardrailMergedEfficiency\% accuracy. We summarize our findings in the Tab.~\ref{tab:guardrail-effectiveness}, where merged adapter refers to the adapter resulting from the SVD-based merging of the LoRA adapters.
